Practical Parametricity for GADTs

### 2020-07-14 20-gadts/inf-rules.tex
small repairs

\[\begin{array}{c}
\AXC{$\Gamma; \Phi \vdash F$}
\AXC{$\Gamma; \Phi \vdash G$}
\AXC{$\Gamma; \Phi \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \emptyset
  \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
\DisplayProof
\\\\
\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
\AXC{$\Gamma; \emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol\psi} \,F \,G$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\psi :=_{\ol\beta} K}]$}
\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\psi :=_{\ol\beta} K}]$}
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi, \ol{\phi} \vdash H$}
\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash F}$}
\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash
    G}$}
\TIC{$\Gamma; \emptyset
  ~|~\emptyset
  \vdash \map^{\ol{F},\ol{G}}_H :
  \Nat^\emptyset\;(\ol{\Nat^{\Phi,\ol{\beta}}\,F\,G})\;
  (\Nat^\Phi\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
      :=_{\ol{\beta}} G}])$} 
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi, \phi, \ol{\alpha} \vdash H$}
\UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
  \Nat^{\Phi,\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
    \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
  \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
\DisplayProof
\\\\
\AXC{$\Gamma; \phi,\Phi,\ol{\alpha} \vdash H$}
\AXC{$\Gamma; \Phi, \ol{\beta} \vdash F$}
\BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
  \Nat^\emptyset\; (\Nat^{\Phi,\ol{\beta}}\,H[\phi
    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\Phi,\ol{\beta}}
    \,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
\AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
\TIC{$\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
  (\Lan^{\ol\alpha}_{\ol K}\,  F)\,\ol{K[\ol{\alpha := A}]}$}  
\DisplayProof
\\\\
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
\AXC{$\ol{\Gamma;\Phi \vdash A}$}
\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol A $}
\TIC{$\Gamma;\Phi~|~\Delta \vdash \partial^{G, \ol K}_F \eta\,t :
  G[\ol{\beta := A}]$}
\DisplayProof
\end{array}\]
\end{dfn}

@@

The fundamental idea underlying Reynolds' parametricity is to give
each type $F(\alpha)$ with one free variable $\alpha$ both an {\em
  object interpretation} $F_0$ taking sets to sets and a
\emph{relational interpretation} $F_1$ taking relations $R :
\rel(A,B)$ to relations $F_1 (R) : \rel(F_0 (A), F_0 (B))$, and to
interpret each term $t(\alpha,x) : F(\alpha)$ with one free term
variable $x : G(\alpha)$ as a map $t_0$ associating to each set $A$ a
homomorphism $t_0(A) : G_0(A) \to F_0(A)$, and to each relation $R$ a
morphism $t_1(R) : G_1(R) \to F_1(R)$.  These interpretations are to
be given inductively on the structures of $F$ and $t$ in such a way
that they imply two fundamental theorems. The first is an
\emph{Identity Extension Lemma}, which states that $F_1(\Eq_A) =
\Eq_{F_0(A)}$, and is the essential property that makes a model
relationally parametric rather than just induced by a logical
relation. The second is an \emph{Abstraction Theorem}, which states
that, for any $R :\rel(A, B)$, $(t_0(A),t_0(B))$ is a morphism in
$\rel$ from $(G_0(A),G_0(B),G_1(R))$ to $(F_0(A),F_0(B),F_1(R))$. The
Identity Extension Lemma is similar to the Abstraction Theorem except
that it holds for {\em all} elements of a type's interpretation, not
just those that are interpretations of terms.  Similar theorems are
expected to hold for types and terms with any number of free
variables.

To accommodate GADTs, we will need to transition Reynolds' approach
from a $\set$-based semantics to a semantics based on
$\omega$-complete partial orders. We denote the category of
$\omega$-complete partial orders ($\oCPO$s) and their sup-preserving
morphisms by $\omega$CPO. The underlying set of an $\oCPO$ $A$ is
denoted $|A|$.  The category $\oCPOR$ of relations on $\omega$CPOs has
as its objects triples $(A,B,R)$, where $A, B: \oCPO$ and $R:
\rel(|A|, |B|)$ and $(\bigvee_{i < \omega} a_i, \bigvee_{i < \omega}
b_i) \in R$ whenever $(a_i)_{i < \omega}$ and $(b_i)_{i < \omega}$ are
chains in $A$ and $B$, respectively, such that $(a_i,b_i) \in R$ for
all $i$. The category $\oCPOR$ has as its morphisms from $(A,B,R)$ to
$(A',B',R')$ pairs $(f : A \to A',g : B \to B')$ of morphisms in
$\oCPO$ such that $(f\,a, g\,b) \in R'$ whenever $(a,b) \in R$. We
note that if $(f,g) : (A,B,R) \to (A',B',R')$ and $(a_i)_{i < \omega}$
and $(b_i)_{i < \omega}$ are chains in $A$ and $B$, respectively, then
$ (f \, (\bigvee_{i < \omega} a_i), g \, (\bigvee_{i < \omega} b_i)) =
(\bigvee_{i < \omega} (f a_i), \bigvee_{i < \omega} (g b_i)) \in R'$
necessarily holds.  We write $R : \oCPOR(A,B)$ in place of $(A,B,R):
\oCPOR$ when convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$
and $\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain}
$B$ of $R$, respectively. If $A : \oCPO$, then we write $\Eq_A =
(A,A,\Eq_{|A|})$ for the {\em equality relation} on $A$.

@@

If $\mcA$ and $\mcC$ are $\omega_1$-cocomplete, then the functor $F :
\mcA \to \mcC$ is {\em $\omega_1$-cocontinuous} if it preserves
$\omega_1$-directed colimits. We write $[\A,\C]_{\omega_1}$ for 
the category of $\omega_1$-cocontinuous functors from $\A$ to $\C$,
and $\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
Since (co)limits of functors are computed pointwise, $C^\A$ has all
(co)limits that $\C$ has, and (co)limits of (co)continuous functors
are again (co)continuous. It follows that $[\A,\C]_{\omega_1}$ is
($\omega_1$1)(co)complete whenever $\C$ is.

If $\mcA$ is locally small, then an object $A$ of $\mcA$ is {\em
  $\omega_1$-presentable} if the functor $\mathsf{Hom}_\mcA(A, -) :
\mcA \to \Set$ preserves $\omega_1$-directed colimits, i.e., if for
every $\omega_1$-directed poset $\mathcal D$ and every functor $F :
{\mathcal D} \to \mcC$, there is a canonical isomorphism $\colim{d \in
  \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq \mathsf{Hom}_{\mcA}(A,
\colim{d \in \mcD}{Fd})$. A locally small category $\mcA$ is {\em
  $\omega_1$-accessible} if it is $\omega_1$-cocomplete and has a set
$\mcA_0$ of $\omega_1$-presentable objects such that every object is
an $\omega_1$-directed colimit of objects in $\mcA_0$; a locally small
category is {\em locally $\omega_1$-presentable} if it is
$\omega_1$-accessible and cocomplete.

The category $\oCPO$ is locally $\omega_1$-presentable (but not
locally finitely presentable); see Examples~1.18(2)
of~\cite{ar94}). Its $\omega_1$-presentable objects are precisely the
$\omega$CPOs cardinality less than $\omega_1$, i.e., the countable
$\omega$CPOs. In the next subsection we will interpret type variables
in $\tvars^k \cup \fvars^k$ as elements of
$[\oCPO^k,\oCPO]_{\omega_1}$; the following special cases of standard
results (see, e.g.,~\cite{ar94}) will therefore be critical to
deducing important properties of our object interpretations of types:
\begin{prop}\label{prop:prelims}
\begin{enumerate}
\item If $\C_1,...,\C_n$ are locally $\omega_1$-presentable categories
  then so is $\C_1 \times ... \times \C_n$. Moreover, the presentable
  objects of $\C_1 \times ... \times \C_n$ are exactly the tuples of
  the form $(P_1,...,P_n)$, where, for each $i = 1,...,n$, the object
  $P_i$ is presentable in $\C_i$.
\item If $\mathcal{A}$ is $\omega_1$-accessible and $\mathcal{C}$ is
  $\lambda$-cocomplete, then the category $[\mathcal{A},
  \mathcal{C}]_{\omega_1}$ is naturally equivalent to the category
  $\mathcal{C}^{\mathcal{A}_0}$.
\item If $\mathcal{C}$ is locally $\omega_1$-presentable and
  $\mathcal{A}_0$ is essentially small, then
  $\mathcal{C}^{\mathcal{A}_0}$ is locally $\omega_1$-presentable.
\end{enumerate}
\end{prop}
\noindent
Together, the statements in Proposition~\ref{prop:prelims} give that
if $\mathcal{A}$ and $\mathcal{C}$ are locally $\omega_1$-presentable,
then $[\mathcal{A}, \mathcal{C}]_{\omega_1}$ is naturally equivalent
to $\mathcal{C}^{\mathcal{A}_0}$, and thus is
$\omega_1$-presentable. Thus, for all $k_1,...,k_n \in \nat^n$,
$[\mathcal{A}, \mathcal{C}]_{\omega_1}^{k_1} \times ... \times
[\mathcal{A}, \mathcal{C}]_{\omega_1}^{k_n}$ is locally
$\omega_1$-presentable, and therefore $[[\mathcal{A},
    \mathcal{C}]_{\omega_1}^{k_1} \times ... \times [\mathcal{A},
    \mathcal{C}]_{\omega_1}^{k_n}, \mathcal{C}]_{\omega_1}$ is as
well. Taking both $\A$ and $C$ to be $\oCPO$ --- as we will to ensure
that the fixpoints interpreting $\mu$-types in $\oCPO$ exist --- we
have
\begin{prop}\label{prop:nat-interps-sets}
  For all $k_1,...,k_n \in \nat^n$,
  \[[[\oCPO, \oCPO]_{\omega_1}^{k_1}
  \times ... \times [\oCPO, \oCPO]_{\omega_1}^{k_n},
  \oCPO]_{\omega_1}\] is locally $\omega_1$-presentable.
\end{prop}

@@

When convenient we identify a functor in $[\oCPO^0, \oCPO]_{\omega_1}$
with its value on $\ast$ and consider a $\oCPO$ environment to map a
type variable of arity $0$ to an $\omega_1$-cocontinuous functor from
$\oCPO^0$ to $\oCPO$, i.e., to an $\oCPO$. If $\Phi =
\{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
$K_i : [\oCPO^{k_i},\oCPO]_{\omega_1}$ for $i = 1,...,n$, then we
write $\rho[\ol{\Phi := K}]$ for the $\oCPO$ environment $\rho'$ such
that $\rho' \phi_i = K_i$ for $i = 1,...,n$ and $\rho' \phi = \rho
\phi$ if $\phi \not \in \Phi$.  If $\rho$ is an $\oCPO$ environment,
we write $\Eq_\rho$ for the $\oCPOR$ environment (see
Definition~\ref{def:reln-env}) such that $\Eq_\rho v = \Eq_{\rho v}$
for every type variable $v$. The categories $\oCPORT_k$ and relational
interpretations appearing in the third clause of
Definition~\ref{def:set-sem} are given in full in
Section~\ref{sec:rel-interp}. We note that $\omega_1$-directed
colimits in $\oCPOenv$ are taken pointwise.

\begin{dfn}\label{def:set-sem}
The {\em object interpretation} $\oCPOsem{\cdot} : \F \to [\oCPOenv,
  \oCPO]_{\omega_1}$ is defined by
\begin{align*}
  \oCPOsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \oCPOsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \oCPOsem{\Gamma; \emptyset
    \vdash \Nat^\Phi
    \,F\,G}\rho &= \{\eta : \lambda \ol{K}. \,\oCPOsem{\Gamma;
    \Phi \vdash
    F}\rho[\ol{\Phi := K}] 
      \Rightarrow \lambda \ol{K}.\,\oCPOsem{\Gamma; 
        \Phi \vdash G}\rho[\ol{\Phi := K}] \\ 
      &\hspace{0.3in}|~\forall \ol{K = (K_0, K_1, K^*) : \oCPORT_k}.\\
      &\hspace{0.4in}(\eta_{\overline{K_0}}, \eta_{\overline{K_1}})
      : \oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]
      \rightarrow \oCPORsem{\Gamma; \Phi \vdash
        G}\Eq_{\rho}[\ol{\Phi := K}] \} \\
  \oCPOsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
  (\rho\phi)\,\ol{\oCPOsem{\Gamma;\Phi \vdash
    F}\rho}\\
  \oCPOsem{\Gamma;\Phi \vdash F+G}\rho &=
  \oCPOsem{\Gamma;\Phi \vdash F}\rho +
  \oCPOsem{\Gamma;\Phi \vdash G}\rho\\
  \oCPOsem{\Gamma;\Phi \vdash F \times G}\rho &=
  \oCPOsem{\Gamma;\Phi \vdash F}\rho \times
  \oCPOsem{\Gamma;\Phi \vdash G}\rho\\ 
  \oCPOsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
    T^\oCPO_{H,\rho})\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}\\
    \text{where } T^\oCPO_{H,\rho}\,F & = \lambda
  \ol{A}. \oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
  \text{and } T^\oCPO_{H,\rho}\,\eta &= \lambda
  \ol{A}. \oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}] \\
  \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho &=
    (\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}
      \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}])
    \ol{\oCPOsem{\Gamma; \Phi \vdash A}\rho}
\end{align*}
\end{dfn}
If $\rho \in \oCPOenv$ and $\vdash F$ then we write $\oCPOsem{\vdash F}$
instead of $\oCPOsem{\vdash F}\rho$ since the environment is
immaterial.

@@

When convenient we identify a $0$-ary $\oCPO$ relation transformer
with the $\oCPO$ relation (transformer) that is its codomain and
consider an $\oCPO$ relation environment to map a type variable of
arity $0$ to an $\oCPO$ relation.  We write $\rho[\ol{\alpha := R}]$
for the $\oCPO$ relation environment $\rho'$ such that $\rho' \alpha_i
\, = R_i$ for $i = 1,...,k$ and $\rho' \alpha = \rho\alpha$ if $\alpha
\not \in \{\alpha_1,...,\alpha_k\}$.  If $\rho$ is an $\oCPO$ relation
environment, we write $\pi_1 \rho$ and $\pi_2 \rho$ for the $\oCPO$
environments mapping each type variable $\phi$ to the functors
$(\rho\phi)^1$ and $(\rho\phi)^2$, respectively.

>>>>

\[\begin{array}{c}
\AXC{$\Gamma; \Phi \vdash F$}
\AXC{$\Gamma; \Phi \vdash G$}
\AXC{$\Gamma; \Phi \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \emptyset
  \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
\DisplayProof
\\\\
\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
\AXC{$\Gamma; \emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol\psi} \,F \,G$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\psi :=_{\ol\beta} K}]$}
\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\psi :=_{\ol\beta} K}]$}
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi, \ol{\phi} \vdash H$}
\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash F}$}
\AXC{$\ol{\Gamma; \Phi,\ol{\beta} \vdash
    G}$}
\TIC{$\Gamma; \emptyset
  ~|~\emptyset
  \vdash \map^{\ol{F},\ol{G}}_H :
  \Nat^\emptyset\;(\ol{\Nat^{\Phi,\ol{\beta}}\,F\,G})\;
  (\Nat^\Phi\,H[\ol{\phi :=_{\ol{\beta}} F}]\;H[\ol{\phi
      :=_{\ol{\beta}} G}])$} 
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi, \phi, \ol{\alpha} \vdash H$}
\UIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \tin_H :
  \Nat^{\Phi,\ol{\beta}} H[\phi :=_{\ol{\beta}} (\mu
    \phi.\lambda \ol{\alpha}.H)\ol{\beta}][\ol{\alpha := \beta}]\,(\mu
  \phi.\lambda \ol{\alpha}.H)\ol{\beta}$}
\DisplayProof
\\\\
\AXC{$\Gamma; \phi,\Phi,\ol{\alpha} \vdash H$}
\AXC{$\Gamma; \Phi, \ol{\beta} \vdash F$}
\BIC{$\Gamma; \emptyset  \,|\, \emptyset \vdash \fold^F_H :
  \Nat^\emptyset\; (\Nat^{\Phi,\ol{\beta}}\,H[\phi
    :=_{\ol{\beta}} F][\ol{\alpha := \beta}]\,F)\; (\Nat^{\Phi,\ol{\beta}}
    \,(\mu \phi.\lambda \ol{\alpha}.H)\ol{\beta}\,F)$}
\DisplayProof
\\\\
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
\AXC{$\ol{\Gamma;\Phi \vdash A} \hspace*{0.3in} \Gamma;\Phi~|~\Delta \vdash t : F[\ol{\alpha :=A}]$}
\TIC{$\Gamma;\Phi~|~\Delta \vdash \int_{\ol K,F} t :
  (\Lan^{\ol\alpha}_{\ol K}\,  F)\,\ol{K[\ol{\alpha := A}]}$}  
\DisplayProof
\\\\
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
\AXC{$\ol{\Gamma;\Phi \vdash B}$}
\AXC{$\Gamma;\Phi~|~\Delta \vdash t : (\Lan^{\ol\alpha}_{\ol K} F)\,\ol B $}
\TIC{$\Gamma;\Phi~|~\Delta \vdash \partial^{G, \ol K}_F \eta\,t :
 G[\ol{\beta := B}]$}
\DisplayProof
\end{array}\]
\end{dfn}

@@

The fundamental idea underlying Reynolds' parametricity is to give
each type $F(\alpha)$ with one free variable $\alpha$ both an {\em
  object interpretation} $F_0$ taking sets to sets and a
\emph{relational interpretation} $F_1$ taking relations $R :
\rel(A,B)$ to relations $F_1 (R) : \rel(F_0 (A), F_0 (B))$, and to
interpret each term $t(\alpha,x) : F(\alpha)$ with one free term
variable $x : G(\alpha)$ as a map $t_0$ associating to each set $A$ a
function $t_0(A) : G_0(A) \to F_0(A)$, and to each relation $R$ a
morphism $t_1(R) : G_1(R) \to F_1(R)$.  These interpretations are to
be given inductively on the structures of $F$ and $t$ in such a way
that they imply two fundamental theorems. The first is an
\emph{Identity Extension Lemma}, which states that $F_1(\Eq_A) =
\Eq_{F_0(A)}$, and is the essential property that makes a model
relationally parametric rather than just induced by a logical
relation. The second is an \emph{Abstraction Theorem}, which states
that, for any $R :\rel(A, B)$, $(t_0(A),t_0(B))$ is a morphism in
$\rel$ from $(G_0(A),G_0(B),G_1(R))$ to $(F_0(A),F_0(B),F_1(R))$. The
Identity Extension Lemma is similar to the Abstraction Theorem except
that it holds for {\em all} elements of a type's interpretation, not
just those that are interpretations of terms.  Similar theorems are
expected to hold for types and terms with any number of free
variables.

To accommodate GADTs, we will need to transition Reynolds' approach
from a $\set$-based semantics to a semantics based on
$\omega$-complete partial orders. We denote the category of
$\omega$-complete partial orders ($\oCPO$s) and their sup-preserving
morphisms by $\omega$CPO. The underlying set of an $\oCPO$ $A$ is
denoted $|A|$.  The category $\oCPOR$ of $\oCPO$ relations has
as its objects triples $(A,B,R)$, where $A, B: \oCPO$ and $R:
\rel(|A|, |B|)$ and $(\bigvee_{i < \omega} a_i, \bigvee_{i < \omega}
b_i) \in R$ whenever $(a_i)_{i < \omega}$ and $(b_i)_{i < \omega}$ are
chains in $A$ and $B$, respectively, such that $(a_i,b_i) \in R$ for
all $i$. The category $\oCPOR$ has as its morphisms from $(A,B,R)$ to
$(A',B',R')$ pairs $(f : A \to A',g : B \to B')$ of morphisms in
$\oCPO$ such that $(f\,a, g\,b) \in R'$ whenever $(a,b) \in R$. We
note that if $(f,g) : (A,B,R) \to (A',B',R')$ and $(a_i)_{i < \omega}$
and $(b_i)_{i < \omega}$ are chains in $A$ and $B$, respectively, then
$ (f \, (\bigvee_{i < \omega} a_i), g \, (\bigvee_{i < \omega} b_i)) =
(\bigvee_{i < \omega} (f a_i), \bigvee_{i < \omega} (g b_i)) \in R'$
necessarily holds.  We write $R : \oCPOR(A,B)$ in place of $(A,B,R):
\oCPOR$ when convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$
and $\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain}
$B$ of $R$, respectively. If $A : \oCPO$, then we write $\Eq_A =
(A,A,\Eq_{|A|})$ for the {\em equality relation} on $A$.

@@

If $\mcA$ and $\mcC$ are $\omega_1$-cocomplete, then a functor $F :
\mcA \to \mcC$ is {\em $\omega_1$-cocontinuous} if it preserves
$\omega_1$-directed colimits. We write $[\A,\C]_{\omega_1}$ for 
the category of $\omega_1$-cocontinuous functors from $\A$ to $\C$,
and $\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
Since (co)limits of functors are computed pointwise, $C^\A$ has all
(co)limits that $\C$ has, and (co)limits of (co)continuous functors
are again (co)continuous. It follows that $[\A,\C]_{\omega_1}$ is
($\omega_1$-)(co)complete whenever $\C$ is.

If $\mcA$ is locally small, then an object $A$ of $\mcA$ is {\em
  $\omega_1$-presentable} if the functor $\mathsf{Hom}_\mcA(A, -) :
\mcA \to \Set$ preserves $\omega_1$-directed colimits, i.e., if for
every $\omega_1$-directed poset $\mathcal D$ and every functor $F :
{\mathcal D} \to \A$, there is a canonical isomorphism $\colim{d \in
  \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq \mathsf{Hom}_{\mcA}(A,
\colim{d \in \mcD}{Fd})$. A locally small category $\mcA$ is {\em
  $\omega_1$-accessible} if it is $\omega_1$-cocomplete and has a set
$\mcA_0$ of $\omega_1$-presentable objects such that every object is
an $\omega_1$-directed colimit of objects in $\mcA_0$; a locally small
category is {\em locally $\omega_1$-presentable} if it is
$\omega_1$-accessible and cocomplete.

The category $\oCPO$ is locally $\omega_1$-presentable (but not
locally finitely presentable); see Examples~1.18(2)
of~\cite{ar94}). Its $\omega_1$-presentable objects are precisely the
$\omega$CPOs cardinality less than $\omega_1$, i.e., the countable
$\omega$CPOs. In the next subsection we will interpret type variables
in $\tvars^k \cup \fvars^k$ as elements of
$[\oCPO^k,\oCPO]_{\omega_1}$; the following special cases of standard
results (see, e.g.,~\cite{ar94}) will therefore be critical to
deducing important properties of our object interpretations of types:
\begin{prop}\label{prop:prelims}
\begin{enumerate}
\item If $\C_1,...,\C_n$ are locally $\omega_1$-presentable categories
  then so is $\C_1 \times ... \times \C_n$. Moreover, the presentable
  objects of $\C_1 \times ... \times \C_n$ are exactly the tuples of
  the form $(P_1,...,P_n)$, where, for each $i = 1,...,n$, the object
  $P_i$ is presentable in $\C_i$.
\item If $\mathcal{A}$ is $\omega_1$-accessible and $\mathcal{C}$ is
  $\lambda$-cocomplete, then the category $[\mathcal{A},
  \mathcal{C}]_{\omega_1}$ is naturally equivalent to the category
  $\mathcal{C}^{\mathcal{A}_0}$.
\item If $\mathcal{C}$ is locally $\omega_1$-presentable and
  $\mathcal{A}_0$ is essentially small, then
  $\mathcal{C}^{\mathcal{A}_0}$ is locally $\omega_1$-presentable.
\end{enumerate}
\end{prop}
\noindent
Together, the statements in Proposition~\ref{prop:prelims} give that
if $\mathcal{A}$ and $\mathcal{C}$ are locally $\omega_1$-presentable,
then $[\mathcal{A}, \mathcal{C}]_{\omega_1}$ is naturally equivalent
to $\mathcal{C}^{\mathcal{A}_0}$, and thus is
$\omega_1$-presentable. Thus, for all $k_1,...,k_n \in \nat^n$,
$[\mathcal{A}^{k_1}, \mathcal{C}]_{\omega_1} \times ... \times
[\mathcal{A}^{k_n}, \mathcal{C}]_{\omega_1}$ is locally
$\omega_1$-presentable, and therefore $[[\mathcal{A}^{k_1},
    \mathcal{C}]_{\omega_1} \times ... \times [\mathcal{A}^{k_n},
    \mathcal{C}]_{\omega_1}, \mathcal{C}]_{\omega_1}$ is as
well. Taking both $\A$ and $C$ to be $\oCPO$ --- as we will to ensure
that the fixpoints interpreting $\mu$-types in $\oCPO$ exist --- we
have
\begin{prop}\label{prop:nat-interps-sets}
  For all $k_1,...,k_n \in \nat^n$,
  \[[[\oCPO^{k_1}, \oCPO]_{\omega_1}
  \times ... \times [\oCPO^{k_n}, \oCPO]_{\omega_1},
  \oCPO]_{\omega_1}\] is locally $\omega_1$-presentable.
\end{prop}

@@

When convenient we identify a functor in $[\oCPO^0, \oCPO]_{\omega_1}$
with its value on $\ast$ and consider a $\oCPO$ environment to map a
type variable of arity $0$ to an $\omega_1$-cocontinuous functor from
$\oCPO^0$ to $\oCPO$, i.e., to an $\oCPO$. If $\Phi =
\{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
$K_i : [\oCPO^{k_i},\oCPO]_{\omega_1}$ for $i = 1,...,n$, then we
write either $\rho[\ol{\Phi := K}]$ or $\rho[\ol{\phi := K}]$ for the
$\oCPO$ environment $\rho'$ such that $\rho' \phi_i = K_i$ for $i =
1,...,n$ and $\rho' \phi = \rho \phi$ if $\phi \not \in \Phi$.  If
$\rho$ is an $\oCPO$ environment, we write $\Eq_\rho$ for the $\oCPOR$
environment (see Definition~\ref{def:reln-env}) such that $\Eq_\rho v
= \Eq_{\rho v}$ for every type variable $v$. The categories
$\oCPORT_k$ and relational interpretations appearing in the third
clause of Definition~\ref{def:set-sem} are given in full in
Section~\ref{sec:rel-interp}. We note that $\omega_1$-directed
colimits in $\oCPOenv$ are taken pointwise.

\begin{dfn}\label{def:set-sem}
The {\em object interpretation} $\oCPOsem{\cdot} : \F \to [\oCPOenv,
  \oCPO]_{\omega_1}$ is defined by
\begin{align*}
  \oCPOsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \oCPOsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \oCPOsem{\Gamma; \emptyset
    \vdash \Nat^\Phi
    \,F\,G}\rho &= \{\eta : \lambda \ol{K}. \,\oCPOsem{\Gamma;
    \Phi \vdash
    F}\rho[\ol{\Phi := K}] 
      \Rightarrow \lambda \ol{K}.\,\oCPOsem{\Gamma; 
        \Phi \vdash G}\rho[\ol{\Phi := K}] \\ 
      &\hspace{0.3in}|~\forall \ol{K = (K^1, K^2, K^*) : \oCPORT_k}.\\
      &\hspace{0.4in}(\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
      : \oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\ol{\Phi := K}]
      \rightarrow \oCPORsem{\Gamma; \Phi \vdash
        G}\Eq_{\rho}[\ol{\Phi := K}] \} \\
  \oCPOsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
  (\rho\phi)\,\ol{\oCPOsem{\Gamma;\Phi \vdash
    F}\rho}\\
  \oCPOsem{\Gamma;\Phi \vdash F+G}\rho &=
  \oCPOsem{\Gamma;\Phi \vdash F}\rho +
  \oCPOsem{\Gamma;\Phi \vdash G}\rho\\
  \oCPOsem{\Gamma;\Phi \vdash F \times G}\rho &=
  \oCPOsem{\Gamma;\Phi \vdash F}\rho \times
  \oCPOsem{\Gamma;\Phi \vdash G}\rho\\ 
  \oCPOsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
    T^\oCPO_{H,\rho})\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}\\
    \text{where } T^\oCPO_{H,\rho}\,F & = \lambda
  \ol{A}. \oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
  \text{and } T^\oCPO_{H,\rho}\,\eta &= \lambda
  \ol{A}. \oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}] \\
  \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho &=
    (\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash K}\rho[\ol{\alpha := \_}]}}
      \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \_}])
    \ol{\oCPOsem{\Gamma; \Phi \vdash A}\rho}
\end{align*}
\end{dfn}
If $\rho \in \oCPOenv$ and $\vdash F$ then we write $\oCPOsem{\vdash F}$
instead of $\oCPOsem{\vdash F}\rho$ since the environment is
immaterial.

@@

When convenient we identify a $0$-ary $\oCPO$ relation transformer
with the $\oCPO$ relation (transformer) that is its codomain and
consider an $\oCPO$ relation environment to map a type variable of
arity $0$ to an $\oCPO$ relation. If $\Phi =
\{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
$K_i : [\oCPOR^{k_i},\oCPOR]_{\omega_1}$ for $i = 1,...,n$, then we
write either $\rho[\ol{\Phi := K}]$ or $\rho[\ol{\phi := K}]$ for the
$\oCPO$ relation environment $\rho'$ such that $\rho' \phi_i = K_i$
for $i = 1,...,n$ and $\rho' \phi = \rho \phi$ if $\phi \not \in
\Phi$. If $\rho$ is an $\oCPO$ relation environment, we write $\pi_1
\rho$ and $\pi_2 \rho$ for the $\oCPO$ relation environments mapping
each type variable $\phi$ to the functors $(\rho\phi)^1$ and
$(\rho\phi)^2$, respectively.
